---
title: "Rolling Stone Data Analysis"
author: "Julia Czekańska"
editor_options: 
  markdown: 
    wrap: sentence
---

## Installing libraries

```{r}
# Define the required libraries
libs <- c("tidyverse", "ggtext", "patchwork", "cowplot", "gridExtra", "lubridate", "ggthemes", "tm", "SnowballC", "wordcloud", "RColorBrewer", "RCurl", "XML", "tidytext", "viridis", "pals", "dplyr","ggplot2", "tidyr", "stringr", "gganimate", "ggraph", "tidygraph")

# Check if libraries are installed and install missing ones
installed_libs <- libs %in% rownames(installed.packages())

if (any(installed_libs == FALSE)) {
    install.packages(libs[!installed_libs])
} else{
  print("All the libraries are already installed")
}

# Load all the needed libraries
for (lib in libs) {
  library(lib, character.only = TRUE)
}
```

## Loading the dataset

The dataset's string values were converted using `str_to_title` due to the case-sensitivity of the queries.

```{r}
df <- read.csv("rolling_stone.csv", stringsAsFactors = FALSE)
string_columns <- c("sort_name", "clean_name", "album", "genre", "type", "spotify_url", "artist_gender")

df[string_columns] <- df[string_columns] %>% lapply(str_to_title)

head(df)
```

## EDA

### Gender Distribution among Artists

Artists were divided into groups and soloists and then amount of male, female, or mixed gender was counted.
The data shows that male artists dominate the rankings and tend to be in bands more often than female artists.

```{r}
gender_distribution <- df %>% 
  mutate(artist_type = ifelse(artist_member_count == 1, "Solo", "Group")) %>%
  count(artist_type, artist_gender) %>%
  filter(!is.na(artist_type)) %>% 
  spread(artist_type, n, fill = 0)
gender_distribution
```

### Artist Member Count vs Debut Album Release Year

The debut years of artists were grouped by decade, and the average number of band members per decade was calculated.
The results highlight the rise of bands in the '60s-'80s.

```{r}
member_vs_debut <- df %>% 
  mutate(debut_decade = floor(debut_album_release_year / 10) * 10) %>%
  group_by(debut_decade) %>%
  summarise(avg_member_count = round(mean(artist_member_count, na.rm = TRUE), 2)) %>% 
  filter(!is.na(debut_decade))
member_vs_debut
```

### Albums with More Than Eight Words in the Title

Album names with more than 8 words were selected, which resulted in displaying 13 names.

```{r}
long_album_titles <- df %>% 
  mutate(word_count = str_count(album, " ") + 1, avg_rank = round(rowMeans(select(., rank_2003, rank_2012, rank_2020), na.rm = TRUE), 2)) %>% 
  filter(word_count > 8) %>% 
  select(album, clean_name, avg_rank) %>% 
  arrange(avg_rank)

long_album_titles
```

### Average Rank Based on Word Usage in Album Title

Album titles were tokenized into individual words, their frequency was counted, and the average rank was calculated for words appearing at least 10 times.
The most common words are generally neutral everyday words.

```{r}
word_rank <- df %>%
  separate_rows(album, sep = " ") %>%
  group_by(album) %>%
  summarise(appearance_count = n(), avg_rank = round(mean(c(rank_2003, rank_2012, rank_2020), na.rm = TRUE), 2)) %>%
  filter(appearance_count >= 10) %>%
  arrange(desc(appearance_count))

word_rank
```

### Top genres by average Spotify popularity

At first rows where genre is NA were removed and then grouped by genre.
Afterwards find the average from column spotify_popularity.
The range is 0 - 100 and the bigger the number, the more popular the album is - so Latin albums are the most popular.

```{r}
df %>%
  filter(!is.na(genre)) %>%
  group_by(genre) %>%
  summarise(avg_popularity = mean(spotify_popularity, na.rm = TRUE)) %>% 
  arrange(desc(avg_popularity))
```

### Number of albums from each genre in each rank

To count the numbers of albums in each rank by genre columns like "rank_2003" were transfered to longer data.
Name of the column transformed to "rank_year" and values were added to column "rank".
After that - genres and rank were filtered, grouped by it and summarized.

```{r}
number_of_albums_from_each_genre_in_each_rank <- df %>%
    pivot_longer(cols = starts_with("rank_"),
               names_to = "rank_year",
               values_to = "rank") %>%
  filter(!is.na(rank) & !is.na(genre)) %>%
  group_by(genre, rank_year) %>% 
  summarise(album_count = n(), .groups = "drop") 

number_of_albums_from_each_genre_in_each_rank
```

### Top 10 albums that improved the most (positive)

Albums that aren't in 2003 or 2020 rankings were filtered out and arranged by *differential* descending to choose top 10.

```{r}
top_improved <- df %>%
  filter(!is.na(rank_2003) & !is.na(rank_2020)) %>%
  arrange(desc(differential)) %>% 
  select(clean_name, album, rank_2003, rank_2020, differential) %>%
  slice_head(n = 10)
top_improved
```

### Top 10 albums that dropped the most (negative)

Albums that aren't in 2003 or 2020 rankings were filtered out and arranged by *differential* ascending to choose top 10.

```{r}
top_dropped <- df %>%
  filter(!is.na(rank_2003) & !is.na(rank_2020)) %>%
  arrange(differential) %>%
  select(clean_name, album, rank_2003, rank_2020, differential) %>%
  slice_head(n = 10)
top_dropped
```

### Percentage of the albums in top charts with special characters in their name

Special characters were defined other than dots and commas, then *grepl* function is used to count how many albums have special characters in their name, divided by the number of distinct names of the albums.

```{r}
special_characters <- "[!@#$%^&*()_+={}|:;\"<>?/~`\\]"

albums_with_special_chars <- df %>%
  filter(grepl(special_characters, album)) %>%
  select(album)

num_special_char_albums <- nrow(albums_with_special_chars)

total_albums <- nrow(df %>% distinct(album))

percentage_special_char_albums <- (num_special_char_albums / total_albums) * 100

num_special_char_albums
percentage_special_char_albums
```

### Albums that placed in top 25 in all three years

Albums that were ranked in top 25 in all three rankings are filtered and selected.

```{r}
df %>%
  filter(rank_2003 <= 25 & rank_2012 <= 25 & rank_2020 <= 25) %>%
  select(clean_name, album, genre, rank_2003, rank_2012, rank_2020)
```

## Plots

### Average age of band members for each genre at the moment of debut

Aaverage age of artists at the moment of debut was calculated and then the data filtered.
It can be noticed that most genres have a central cluster around 20-30 years, which suggests that most bands release albums in their 20s.
The results also show that genres like Soul/Gospel/R&B show a wider spread that indicates bands of different ages releasing albums.

```{r}
rolling_stone <- df %>%
  mutate(Band.Age = release_year - (artist_birth_year_sum / artist_member_count)) %>%
  filter(!is.na(Band.Age) & Band.Age > 0 & artist_member_count > 1 & !is.na(genre)) %>% 
  group_by(genre) %>%
  filter(n() >= 2)

ggplot(rolling_stone, aes(x = genre, y = Band.Age, fill = genre)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  scale_fill_manual(values = stepped(length(unique(rolling_stone$genre)))) +
  labs(
    title = "Band Age Distribution at Album Release by Genre",
    x = "Genre",
    y = "Band Age at Album Release"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

### Heatmap of number of albums from each genre in each rank

This plot uses previously calculated *number_of_albums_from_each_genre_in_each_rank* and vizualizes it as a heatmap.

```{r}
selected_colors <- c(stepped(20)[3], stepped(20)[11], stepped(20)[19])


ggplot(number_of_albums_from_each_genre_in_each_rank,
       aes(x = rank_year, y = genre, fill = album_count)) +
  
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradientn(
    colors = selected_colors,
    name = "Album Count",
    breaks = seq(0, max(number_of_albums_from_each_genre_in_each_rank$album_count) - 10, by = 10)
  ) +
  theme_minimal() +
  labs(title = "Number of Albums in Ranking",
       x = "Ranking Year",
       y = "Genre") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        legend.position = "right")
```

### Word cloud of the most used words

The names of the albums are selected, tokenized, and deprived of common "stop words. Afterwards, *worldcloud* function is used, which does the vizualization. The most common word is love.

```{r}
album_words <- df %>%
  select(album) %>% 
  tidytext::unnest_tokens(word, album) %>%  
  anti_join(stop_words, by = "word") %>% 
  count(word, sort = TRUE)

album_words %>%
  arrange(desc(n)) %>% print()

selected_colors <- c(stepped(20)[9], stepped(20)[10], stepped(20)[13], stepped(20)[14], stepped(20)[17], stepped(20)[18])

wordcloud(
  words = album_words$word, 
  freq = album_words$n, 
  min.freq = 4,
  max.words = 50,
  random.order = FALSE,
  colors = selected_colors
)
```

### Average Age by Genre and Artist Gender in 2003

The data is filtered and the average *birth_year* is calculated and grouped by genre and artist gender.
For some genres there aren't any female artists or groups that have both female and male members.
It can be noticed that the age of the female artists is lower in most genres.

```{r}
average_birthyear_by_genre_gender <- df %>%
  filter(!is.na(artist_gender) & !is.na(genre)) %>%
  mutate(birth_year = ifelse(artist_member_count > 1,
                             artist_birth_year_sum / artist_member_count,
                             artist_birth_year_sum)) %>% 
  filter(!is.na(artist_gender) & !is.na(birth_year)) %>% 
  mutate(age = (2003 - birth_year)) %>%  
  group_by(genre, artist_gender) %>%
  summarise(average_birthyear = mean(birth_year, na.rm = TRUE), 
            average_age = mean(age, na.rm = TRUE),  
            .groups = "drop")

selected_colors <- c(stepped(20)[3], stepped(20)[15], stepped(20)[11])

ggplot(average_birthyear_by_genre_gender, aes(x = genre, y = average_age, color = artist_gender, group = artist_gender)) +
  geom_line(linewidth = 1) + 
  geom_point(size = 4) +
  labs(title = "Average Age by Genre and Artist Gender in 2003",
       x = "Genre",
       y = "Average Age",
       color = "Artist Gender") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_line(color = "gray80")) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, by = 10)) +
  scale_color_manual(values = selected_colors)
```

### Correlation Between Average Birth Year and Rank by Year and Genre

Each artist’s average birth year is computed.
The code reshapes their chart‑rank data into a long format by year, and then produces a faceted scatterplot showing the relationship between rank and average birth year colored by genre.
We can see that a lot of artists were born in 1940-1950 years, which makes sense, since there was a "baby boom" after the II World War.

```{r}
debut_age_rank_data <- df %>%
  filter(!is.na(artist_gender) & !is.na(genre)) %>%
  mutate(birth_year = ifelse(artist_member_count > 1,
                             artist_birth_year_sum / artist_member_count,
                             artist_birth_year_sum)) %>%
  pivot_longer(cols = starts_with("rank_"),
               names_to = "rank_year",
               values_to = "rank") %>%
  filter(!is.na(birth_year) & !is.na(rank)) %>% 
  mutate(rank_year = gsub("rank_", "", rank_year))

plot <- ggplot(debut_age_rank_data, aes(x = rank, y = birth_year)) +
  geom_point(alpha = 0.7, size = 3, aes(color = genre)) +
  labs(title = "Correlation Between Average Birth Year and Rank by Year and Genre",
       x = "Rank",
       y = "Average Birth Year",
       color = "Genre") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~rank_year, scales = "fixed") + 
  scale_color_manual(values = stepped(15)) +
  scale_y_continuous(limits = c(1910, 2000), 
                     breaks = seq(1910, 2000, by = 10)) +  
  scale_x_continuous(breaks = seq(0, 500, by = 100)) +
  theme(strip.text = element_text(size = 12, face = "bold"))

plot
```

### Spotify Presence by Genre

An additional column *has_spotify* has been created, the data was grouped and summarized.

```{r}
spotify_analysis <- df %>%
  filter(!is.na(genre)) %>% 
  mutate(has_spotify = ifelse(!is.na(spotify_url), "Has Spotify", "No Spotify")) %>%
  group_by(genre, has_spotify) %>%
  summarise(album_count = n(), .groups = "drop") %>%
  mutate(percent = album_count / sum(album_count) * 100, .by = genre)


selected_colors <- c(stepped(20)[19], stepped(20)[20])
plot <- ggplot(spotify_analysis, aes(x = 2, y = percent, fill = has_spotify)) +
  geom_bar(stat = "identity") +
  coord_polar(theta = "y") +
  geom_text(data = spotify_analysis %>% filter(has_spotify == "No Spotify" & percent > 0),
            aes(label = paste0(round(percent, 1), "%"), y = percent / 2), 
            size = 4) +
  facet_wrap(~genre, nrow = 3) +
  labs(title = "Spotify Presence by Genre", fill = "Spotify Status") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(size = 10, face = "bold"),
        plot.title = element_text(hjust = 0.5, size = 14)) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = selected_colors)
plot

ggsave("larger_graph.jpg", plot = plot, width = 16, height = 10, dpi = 300)
```

### Average Debut Age of Group vs. Solo Artists Across Genres

New column Solo/Group was added which showed that in Latin the average age for groups is the highest one compared to all the other genres and artist types.

```{r}
data <- df %>% 
  filter(!is.na(genre) & !is.na(artist_gender)) %>% 
  mutate(birth_year = ifelse(artist_member_count > 1,
                             artist_birth_year_sum / artist_member_count,
                             artist_birth_year_sum)) %>% 
  mutate(debut_age = debut_album_release_year - birth_year) %>% 
  mutate(artist_type = ifelse(artist_member_count > 1, "Group", "Solo")) %>% 
  group_by(genre, artist_type) %>% 
  summarise(avg_debut_age = mean(debut_age, na.rm = TRUE), , .groups = "drop") 

selected_colors <- c(stepped(20)[16], stepped(20)[8])

plot <- ggplot(data, aes(x = genre, y = avg_debut_age, group = artist_type, color = artist_type)) +
  geom_line(linewidth = 1) +  
  geom_point(size = 3) +  
  labs(title = "Average Debut Age of Group vs. Solo Artists Across Genres", 
       x = "Genre", y = "Average Debut Age", color = "Artist Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        panel.grid.minor = element_blank(), 
        legend.position = "top") +
  scale_y_continuous(limits = c(15, 50), 
                     breaks = seq(15, 50, by = 5)) +
  scale_color_manual(values = selected_colors)

plot 
```

### Album Rank Changes Over Time

The data is reshaped and the missing values are filtered out.
After that, ranks over time are compared.
An animated plot showing the rank changes for each album, with colors representing the change and shapes showing the genre is displayed it the *Viewer*.

```{r}
data <- df %>%
  pivot_longer(cols = starts_with("rank_"),
               names_to = "rank_year",
               values_to = "rank") %>% 
  filter(!is.na(rank) & !is.na(genre)) %>%
  mutate(rank_year = str_extract(rank_year, "\\d{4}")) %>% 
  mutate(rank_year = as.integer(rank_year)) %>% 
  arrange(rank_year, rank)

data <- data %>%
  group_by(album) %>%
  arrange(rank_year) %>%
  mutate(rank_change = rank - lag(rank, default = rank[1])) 


plot <- ggplot(data, aes(x = album, y = rank, color = rank_change, shape = genre)) +
  geom_point(size = 3) +
  scale_color_gradient2(low = stepped(24)[1], mid = stepped(24)[24], high = stepped(24)[13], midpoint = 0) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14)) + 
  labs(title = "Album Rank Changes Over Time: {frame_time}", 
       x = "Album", y = "Rank", color = "Rank Change", shape = "Genre") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 14),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        ) +
  transition_time(rank_year) +  
  ease_aes('linear')  

animate(plot, fps = 10, duration = 8, width = 900, height = 600, renderer = gifski_renderer())
```

### Appearance of Emotional Words in Album Titles

Emotional words in album titles were counted and categorized as positive or negative.
The visualization shows that positive words appear more frequently, with "love" being the most common.

```{r}
positive_words <- c("Love", "Like", "Happy", "Joy", "Smile", "Dream", "Hope", "Bright", "Peace", "Sweet")
negative_words <- c("Hate", "Dislike", "Sad", "Cry", "Tears", "Dark", "Pain", "Lonely", "Broken", "Fear")

emotional_words <- c(positive_words, negative_words)

emotion_analysis <- df %>%
  separate_rows(album, sep = " ") %>%
  filter(album %in% emotional_words) %>%
  group_by(album) %>%
  summarise(appearance_count = n()) %>%
  arrange(desc(appearance_count))

emotion_analysis <- emotion_analysis %>%
  mutate(category = ifelse(album %in% positive_words, "Positive", "Negative"))

emotion_summary <- emotion_analysis %>%
  group_by(category) %>%
  summarise(total_appearances = sum(appearance_count))

emotion_plot <- ggplot(emotion_analysis, aes(x=reorder(album, -appearance_count), y=appearance_count, fill=category)) +
  geom_col() +
  theme_minimal() +
  scale_fill_manual(values = stepped(24)[c(4, 20)]) +
  labs(title="Emotional Words in Album Titles", x="Word", y="Count") +
  theme(axis.text.x = element_text(angle=45, hjust=1))
emotion_plot
```

### Artist Gender by Birth Decade

An approximated decade of birth was first calculated for each artist, then occurrences were counted and displayed.
Only in recent years more top female than top male artists were born.
Also, most artists in the rankings were born in the '40s-'60s range.

```{r}
rolling_stone <- df %>%
  mutate(birth_decade = floor(artist_birth_year_sum / artist_member_count / 10) * 10) %>%
  filter(!is.na(birth_decade))

gender_by_decade <- rolling_stone %>%
  count(birth_decade, artist_gender) %>%
  spread(artist_gender, n, fill = 0)

g <- ggplot(gather(gender_by_decade, Gender, Count, -birth_decade), aes(x=birth_decade, y=Count, fill=Gender)) +
  geom_bar(stat='identity', position='dodge') +
  theme_minimal() +
  scale_fill_manual(values = stepped(24)[c(12, 16, 20)]) +
  scale_x_continuous(breaks = seq(min(1900), max(2000), by = 10)) +
  labs(title="Artist Gender Distribution by Birth Decade", x="Birth Decade", y="Count")
g
```

### Years Between Debut and Top Album in 2020

Average rank was calculated based on years between debut and 2020.
The plot suggests that artists with shorter careers achieve better rankings.

```{r}
df <- df %>% 
  mutate(years_between_top_2020 = release_year - debut_album_release_year,
         years_range = floor(years_between_top_2020 / 10) * 10,
         years_range_label = paste0(years_range, "-", years_range + 9)) %>%
  filter(!is.na(debut_album_release_year))

years_between_top_2020_range <- df %>%
  select(clean_name, years_between_top_2020, rank_2020, years_range_label) %>%
  filter(!is.na(rank_2020)) %>%
  group_by(years_range_label) %>%
  summarise(avg_rank = round(mean(rank_2020, na.rm = TRUE), 2)) %>%
  arrange(years_range_label)

ggplot(years_between_top_2020_range, aes(x = years_range_label, y = avg_rank, group = 1, color = years_range_label)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = stepped(24)[c(4, 8, 12, 16, 20, 24)]) +
  labs(title = "Average Rank vs. Years Between Debut and 2020 Hit",
       x = "Years Between Debut and 2020 Hit (Range)",
       y = "Average Rank",
       color = "Years Range") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Album Ranking Movements Over Time

This bar plot created by calculating average movement of ranks by release decade shows that newer albums tend to achieve higher places easier.

```{r}
average_movement_by_decade <- df %>% 
  mutate(release_decade = floor(release_year / 10) * 10) %>%
  group_by(release_decade) %>%
  summarise(avg_movement = round(mean(differential, na.rm = TRUE), 2))

ggplot(average_movement_by_decade, aes(x = factor(release_decade), y = avg_movement, fill = factor(release_decade))) +
  geom_col() +
  scale_fill_manual(values = stepped(24)[c(1, 4, 9, 12, 13, 16, 17)]) +
  theme_minimal() +
  labs(title = "Average Album Ranking Movement Over Time", x = "Release Decade", y = "Average Movement", fill = "Decade") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Average Length of Album Titles by Gender

An animated bar plot shows differences in album title length and rank divided by artist gender.
The result shows that for male artists the title length has little influence over rank, for female artists shorter titles are better, and for mixed groups - there is only a small difference in length between the top and bottom ranks.

```{r}
album_length <- df %>% 
  mutate(word_count = str_count(album, " ") + 1, char_count = nchar(album), rank_group = cut(rank_2020, breaks = seq(0, 500, 100))) %>% 
  drop_na(rank_2020)

animated_plot <- ggplot(album_length, aes(x = rank_group)) +
  geom_bar(aes(y = char_count, fill = "Character Count"), stat = "summary", fun = "mean", position = "dodge") +
  geom_bar(aes(y = word_count, fill = "Word Count"), stat = "summary", fun = "mean", position = "dodge") +
  scale_fill_manual(values = stepped(2)) +
  theme_minimal() +
  labs(title = "Album Title Length by Rank Group - {closest_state}", 
       x = "Rank Group", 
       y = "Average Count", 
       fill = "Legend") +
  transition_states(artist_gender)

animated_plot
```

### Relationship Between Debut Album Release Year, Release Year, and Highest Rank

A scatter plot was created to show the relationship between an artist's debut album release year and the number of years it took them to reach their top-ranked album.
The color gradient represents the highest rank achieved, with blue indicating better ranks and red indicating worse ranks.
The results suggest that artists debuting earlier generally took longer to reach their highest-ranked album, but there are exceptions across different time periods.

```{r}
debut_release_rank <- df %>% 
  mutate(highest_rank = pmin(rank_2003, rank_2012, rank_2020, na.rm = TRUE),
         year_difference = release_year - debut_album_release_year) %>% 
  select(debut_album_release_year, release_year, year_difference, highest_rank) %>% 
  arrange(highest_rank)

ggplot(debut_release_rank, aes(x = debut_album_release_year, y = year_difference, color = highest_rank)) +
  geom_point(alpha = 0.6) +
  scale_color_gradientn(colors = stepped(24)[c(17, 13, 9, 1)]) +
  scale_x_continuous(breaks = seq(min(1930), max(2020), by = 10)) +
  theme_minimal() +
  labs(title = "Debut Year vs. Years to Top Album vs. Highest Rank",
       x = "Debut Album Release Year",
       y = "Difference in Years",
       color = "Highest Rank") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(10, 10, 10, 10)
  )
```

### Genres and Billboard Longevity

Singer-Songwriter/Heartland Rock has the highest median value, which tells us that albums in this genre tend to stay on Billboard the longest.
The opposite of this insight applies to Reggae.
Also, genres like Indie/Alternative Rock have relatively wider distributions, which indicates variability in how long albums from these genres stay on the Billboard.

```{r}
df_clean <- df %>% filter(!is.na(weeks_on_billboard) & !is.na(genre))

ggplot(df_clean, aes(x = reorder(genre, weeks_on_billboard, median), y = weeks_on_billboard, fill = genre)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Genres and Billboard Longevity",
       x = "Genre",
       y = "Weeks on Billboard") +
  theme(legend.position = "none") +
  scale_fill_manual(values = stepped(n_distinct(df_clean$genre)))
```

### Soloist Age and Billboard Success

The resulting plot shows that artists that reach Billboard are mostly around 20-30 years old.
It also shows that older men have a higher change of achieving it than older women.

```{r}
df_solo <- df %>% filter(artist_member_count == 1 & !is.na(artist_gender))

ggplot(df_solo, aes(x = artist_gender, y = ave_age_at_top_500, fill = artist_gender)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_minimal() +
  labs(title = "Soloist Age and Billboard Success",
       x = "Artist Gender",
       y = "Average Age at Top 500") +
  scale_fill_manual(values = stepped(2)) +
  theme(legend.position = "none")+
  scale_y_continuous(breaks = seq(0, max(df_solo$ave_age_at_top_500), by = 10))
```

### Genre vs. Rank in a Particular Year

A heatmap was created to visualize the rankings of different genres over multiple years.
Higher rankings (closer to 1) are shown in blue, while lower rankings are in red.
As can be seen, Reggae and Funk/Disco were present near the top of the ranking for all three lists.

```{r}
df_long <- df %>% 
  select(genre, rank_2003, rank_2012, rank_2020) %>% 
  gather(year, rank, -genre) %>% 
  mutate(year = as.factor(gsub("rank_", "", year))) %>% 
  drop_na()

heatmap_plot <- ggplot(df_long, aes(x=year, y=genre, fill=rank)) +
  geom_tile() +
  scale_fill_gradientn(colors = stepped(24)[c(20, 16, 12, 8, 4)], trans = "reverse") +
  theme_minimal() +
  labs(title="Heatmap of Genres vs. Rank in Particular Years", x="Year", y="Genre", fill="Rank")

heatmap_plot
```

### Genre Proportions Over Decades

This plot shows how popularity of genres changes by decades based on their appearance in the rankings.
The results show that in the '70s Big Band/Jazz was most popular, with Hip-Hop/Rap taking its place in the recent decades.

```{r}
df$release_decade <- floor(df$release_year / 10) * 10

genre_proportion <- df %>% 
  count(release_decade, genre) %>% 
  group_by(release_decade) %>% 
  mutate(proportion = n / sum(n)) %>%
  filter(!is.na(genre))

proportion_plot <- ggplot(genre_proportion, aes(x=release_decade, y=proportion, color=genre, group=genre)) +
  geom_line(size=1) +
  geom_point(size=2) +
  scale_color_manual(values = stepped(length(unique(genre_proportion$genre)))) +
  scale_x_continuous(breaks = seq(min(genre_proportion$release_decade), max(genre_proportion$release_decade), by = 10)) +
  theme_minimal() +
  labs(title="Genre Proportions Over Decades",
       x="Release Decade",
       y="Proportion",
       color="Genre")

proportion_plot
```

### Name Length vs. Top Rank vs. Debut Year

This scatter plot examines the relationship between artist name length, debut year, and highest ranking, using color to indicate debut year trends.
As it can be seen, the average name length varies in the 9-13 character amount range and the longest names mostly belong in the '60s-'80s range.

```{r}
df <- df %>% 
  mutate(name_length = nchar(clean_name),
         highest_rank = pmin(rank_2003, rank_2012, rank_2020, na.rm = TRUE)) %>%
  drop_na(name_length, highest_rank, debut_album_release_year)

name_length_plot <- ggplot(df, aes(x=name_length, y=highest_rank, color=debut_album_release_year)) +
  geom_point(alpha=0.6) +
  scale_color_gradientn(colors = stepped(24)[c(1,5,9,13,17)]) +
  theme_minimal() +
  labs(title="Name Length vs. Top Rank vs. Debut Year", x="Name Length", y="Top Rank", color="Debut Year")

name_length_plot
```

### Top 2 artists per genre based on album rankings

This circular plot based on tidytuesday's Lisa's Garden shows top 2 artists from each genre in a colorful way.

```{r}
data_clean <- df %>%
  select(clean_name, genre, rank_2003, rank_2012, rank_2020) %>%
  pivot_longer(cols = starts_with("rank_"), names_to = "year", values_to = "rank") %>%
  filter(!is.na(rank)) %>%
  filter(!is.na(genre)) %>%
  filter(!is.na(clean_name))

top_artists <- data_clean %>%
  group_by(genre, clean_name) %>%
  summarise(best_rank = min(rank, na.rm = TRUE), .groups = "drop") %>%
  arrange(genre, best_rank) %>%
  group_by(genre) %>%
  slice_min(order_by = best_rank, n = 2, with_ties = FALSE)

edges_level0_1 <- top_artists %>%
  distinct(genre) %>%
  mutate(from = "Top Artists", to = genre)

edges_level1_2 <- top_artists %>%
  select(from = genre, to = clean_name)

edge_list <- bind_rows(edges_level0_1, edges_level1_2)

graph <- as_tbl_graph(edge_list, directed = TRUE) %>%
  activate(nodes) %>%
  mutate(genre = ifelse(name %in% top_artists$clean_name, 
                        top_artists$genre[match(name, top_artists$clean_name)], 
                        name))

color_palette_graph <- stepped(length(unique(top_artists$genre)) + 1)

p <- ggraph(graph, layout = 'dendrogram', circular = FALSE) + 
  geom_edge_diagonal() +
  geom_node_point(aes(color = factor(genre)), na.rm = TRUE) +
    geom_node_text(aes(filter = leaf, label = name), repel = TRUE, na.rm = TRUE) +
  scale_color_manual(values = color_palette_graph) +
  theme_void()

node_positions <- ggplot_build(p)$data[[2]]

ggraph(graph, layout = 'dendrogram', circular = TRUE) +
  geom_edge_diagonal(color = "grey60") +
  geom_node_point(aes(color = factor(genre)), size = 3, na.rm = TRUE) +
  geom_node_text(aes(filter = leaf, label = name), size = 3,repel = TRUE, na.rm = TRUE) +
  scale_color_manual(name = "Genres", values = color_palette_graph) +
  theme_void() +
  ggtitle("Top 2 Artists per Genre")
```
